Comment: Inferring Trade Costs from Trade Booms and Trade Busts

### 2016-07-28 Paper/Manuscript.tex
Corrections texte
Mes corrections du 28 juillet 2016

\author[1]{Guillaume Corlay}
\author[2]{Stéphane Dupraz}
\author[4]{Claire Labonne}
\author[1]{Anne Muller}
\author[3]{Guillaume Daudin\footnote{Corresponding author: guillaume.daudin@dauphine.fr, }\footnote{We are grateful to Celine Antonin (OFCE) for her constant supply of helpful comments. She, of course, bears no responsibility for remaining errors.}}
\affil[1] {ENSAE, French National School of Statistics and Economic Administration, 3, avenue Pierre Larousse, 92245 Malakoff, FRANCE}
\affil[2] {Columbia University, 10025 New York NY, USA}
\affil[3] {Université Paris-Dauphine, PSL Research University, LEDa, 75016 PARIS, FRANCE
Université Paris-Dauphine, PSL Research University, LEDa, UMR [225], DIAL, 75016 PARIS, FRANCE

@@

\affil[4] { Paris School of Economics / Université Paris 1 Panthéon Sorbonne -- ACPR - Banque de France, 75009, PARIS, FRANCE}

@@

\begin{abstract}
\cite{JMN2011} offer an alternative to price gaps to quantify
trade costs. Their method consists in deducing international
trade costs from trade flows. They argue that the reduction in
trade costs was the main driving force of the growth of trade
during the first globalization (1870-1913), whereas economic expansion was the main driving force during the second globalization (1950-2000). We argue
that this important result is  driven by the use of an ad hoc aggregation
method. What \cite{JMN2011} capture is the 
difference in the relative starting trade of dyads experiencing faster trade growth in the first and second globalization.
More generally, we cast doubts on the possibility to reach
conclusions of such nature with a method that infers trade
costs from trade flows, and then uses these costs to explain
trade flows. It can only rephrase the information already
contained in openness ratios.
\end{abstract}

@@

Jacks and his coauthors offer in several papers  a method for
inferring international trade costs from trade flows.\footnote{The method is developed in\cite{JMN2008}, \cite{JMN2010}, \cite{NOVY} and \cite{JMN2011}. We will use this latter paper as a reference} Using the general equilibrium model of
\cite{AW2003}, they calculate trade costs (defined as as all barriers to trade, notably transport and transaction costs) and their
evolution during the first and second waves of globalization
(1870-1913 and 1950-2000) as well as the interwar period
(1921-1939) thanks to the impressive set of data they collected
on trade flows, GDP and exports between 27
countries\footnote{Argentina, Australia, Austria, Belgium,
Brazil, Canada, Denmark, France, Germany, Greece, India,
Indonesia, Italy, Japan, Mexico, the Netherlands, New Zealand,
Norway, the Philippines, Portugal, Spain, Sri Lanka, Sweden,
Switzerland, the United Kingdom, the United States, and
Uruguay. The data contain 130 country pairs.}. They provide a
decomposition of the growth of trade caused by the reduction in
trade costs and economic expansion. They use their computations to underline
a difference of nature between the two globalizations:

@@

This is potentially an important result that sheds light on the globalization processes. However, that result is actually driven by an ad hoc
method of aggregation that captures structure effects. The inferred difference between the two globalization comes from the fact that the dyads with the 
fastest growing trade in the first wave of globalization start with very small 
trade. This is not the case in the second wave of globalization.
Using a three-countries version of \cite{AW2003}'s model for
the aggregation of bilateral trade costs shows that there was no 
difference in nature betwee the two globalizations. More generally, we
cast doubt on the possibility to distinguish the effect of trade costs and economic expansion through an approach that relies only on the study of trade
flows. Deducing trade costs from trade flows, and then using
trade costs to explain trade flows is essentially a circular
reasoning. Therefore, Jacks et al.'s approach cannot be an
alternative to traditional investigations of impediments to
trade, such as commodity price gaps.

This paper starts by presenting Jacks et al.'s approach to the
measure of trade costs, and insists on its relevance at the
bilateral level to control for multilateral trade barriers in
gravity regressions (section \ref{1}). It then highlights how paradoxical the
result on a difference of nature between the two globalizations
is since it cannot be deducted from a comparison of the
evolution of openness ratios. Section \ref{2} shows that the
conclusion is only driven by the authors' ad hoc aggregation
method. We propose a microfounded way to aggregate trade costs
and the puzzle fades away. Section \ref{3} explores the reasons
why Jacks et al.'s aggregation technique ends up providing
different results for the two globalizations. We argue that
what Jacks et al. misleadingly attribute to unequal trade costs
decreases between the two globalizations is instead a
difference in the distribution of trade growth over trading
partners.

@@

\cite{AW2003} use this model to microfound gravity equations
and solve \cite{MAC}'s border puzzle by highlighting that
bilateral trade does not depend on bilateral trade barriers per
se, but bilateral trade barriers relative to impediments to
trade with all other trading partners. \cite{AW2003} show that
equilibrium imposes the following relation:

@@

where $x_{ij}$ are real exports from $i$ to $j$, $y_i$ is real
output of country $i$, $y^W$ is the world real output, $\sigma$
is the elasticity of substitution, $P_j$ is the price index in
country $j$ and can be interpreted as multilateral trade
barriers, and $t_{ij}$ is the trade costs factor between $i$
and $j$. Trade costs factors are assumed to be symmetric, i.e.
$t_{ij}$ = $t_{ji}$.

Jacks et al. first depart from \cite{AW2003} by eliminating the
multilateral resistance variable from the gravity equation.
Following \cite{NOVY}, bilateral trade barriers are not
considered relative to multilateral trade barriers modelized by
the price index, but to domestic trade costs. In this case,
trade flows are no longer compared to outputs, but to internal
trade $x_{ii}$.\footnote{Due to data limitations, the authors
use the relation $x_{ii}=GDP_i-EXPORTS_i$ to get internal
trade. We follow them. Concerns about the fact that GDP is
measured in added-value and exportations as gross value are
addressed in appendix B of \cite{JMN2011}.} The equation above
then becomes:

@@

Where the last equality defines $\tau_{ij}$, the geometric mean
of trade costs from country $i$ to $j$ and $j$ to $i$
\emph{relative to intra-national trade costs in countries $i$
and $j$} (no assumption of symmetry of bilateral trade costs is
imposed). It is the trade cost measure used by the authors.

This simple step offers a significant breakthrough in the
microfoundations of gravity equations. Comparing bilateral
trade flows to intranational trade allows to integrate
\cite{AW2003}'s caveat against omitted variable bias while
getting rid of multilateral trade barriers. This is important
when multilateral trade barriers cannot be estimated, e.g. when
we have only data on trade of one country. \cite{JMN2011} take
advantage of this feature in section 5 of their article in
regressing their measure of trade costs between two countries
on a set of proxies for trade costs, such as the distance
between trade partners, tariffs, or the volatility of the
exchange rate. This is a very useful approach.\\

@@

Our point is that a method that use trade flows
alone can teach us much about their determinants. When
departing from \cite{AW2003}'s multilateral resistances, the
arbitrage condition Jacks et al. base their computation on is an
equality between a Marginal Rate of Substitution (MRS) and a
price ratio for a CES utility function, or more precisely the
product of two such equalities, one from the program of each
trade partner. Appendix \ref{A1} derives equation (\ref{keyeq})
this way without passing through multilateral resistance. The
inference of trade costs from trade flows is tantamount to
using the relation between quantities and prices given by a
demand curve derived from fully specified exogenous
preferences.

This is of course no reason in itself to disregard the method.
But when it cannot be used to answer \emph{``the central
question of what drive trade booms and busts''}(p. 186) as the
authors do in section 6 of their article. That is because the reasoning is
circular: first calculate trade costs from trade flows, then
explain trade flows from the inferred trade costs.

Formally, the authors take the logarithm of the key-equation of
their article (\ref{keyeq}) to decompose the product of
bilateral trade flows between $i$ and $j$ in four terms:

@@

Since the trade costs measure is assimilated to trade flows
relative to domestic activity, they simply restate
 an information already contained in any measure of
trade flows relative to economic size, like an openness ratio
defined for instance as the ratio of exportations to GDP.
However the main conclusion in Jacks et al.'s article, namely
that the first globalization was driven more by the decrease in
trade costs than the second one requires the absence of circularity.
Indeed, when we decompose the level of
exportations of a country as the product of its GDP and its
openness ratio, we find results that are different from theirs (see table \ref{OR}).\footnote{We follow
the authors in interpreting log differences as percentages. However, one should keep in mind that given the
size of the changes, this is a very inexact approximation.} The increase in exportations is mainly explained by the growth of
GDP for the large majority of the 27 countries present in the
data. On average, this decomposition attributes 74\% (183/246) of
the growth in trade to the increase in GDP in the second
globalization and 60\% in the first globalization. This is very much in contrast to 
Jacks et al. :\footnote{Contributions of growth in income
similarity and of change in multilateral factors are negative.}

@@

This contrast between both decompositions of the growth of exports is at the center of the argument by Jack et al. As the next section shows, it is actually driven by the ad hoc way they aggregate trade costs.

\section{\label{2} Ad hoc aggregation of trade costs}

Jacks et al.'s conclusion on a difference of nature between the
two globalizations is based on an aggregate trade costs
measure that provides a summary statistic of the
evolution of trade costs across all dyads in the sample, as
well as for different regions of the world. To move from
bilateral costs $\tau_{ij}$ to an aggregate measure of trade
costs, the authors use an arithmetic mean over dyads, weighted
by the sum of the GDP of the two trade partners.

@@

\cite{AW2003}'s model can be used to calculate the aggregate
trade costs a single country is facing in its trade with all
its trading partners. Indeed, using a two-countries model where
all trade partners of country $i$ are treated as one single
country provides an aggregate measure of trade costs that does
not rely on any ad hoc aggregation method but instead on
microfoundations. Since not all trade flows data are available,
it is also possible to use a three-countries model, with
country $i$, its trading partners present in the data, and the
rest of the world.

Let us note $\upsilon_{i}$ the corresponding trade cost faced
by country $i$ with all its trading partners in the data. Its
expression is given by Jacks et al.'s key equation
(\ref{keyeq}) applied to a three-countries model:

@@

Using this measure of trade costs, we provide the same
decomposition in four terms of the increase in trade flows
during the two globalizations as in the authors' article for
the countries with the most trade partners available in the
data (France (24), the UK (25), the USA (23)). We also provide
the unweighted and end-of-period-GPD-weighted averages for all
countries in the sample. Results are displayed in tables
\ref{glob1} and \ref{glob2} along with the results with the
Jacks et al's method.\footnote{They include unweighted and
end-of-period-GPD-weighted averages over country dyads of
bilateral trade costs, such as provided in \cite{JMN2011}. We
also display for France, the UK and the USA the trade cost
measure averaged over trading partners faced by one country, as
the results are presented (with a smaller data set) in
\cite{JMN2008, JMN2010}. We also provide both averages of this
measure over all countries in the data set.}\\

\begin{table}
\centering
\begin{tabular}{|l|c|c|c|c|c|}
\hline
\vbox{\hbox{\strut 1870-1913}\hbox{\strut }\hbox{\strut}} & \vbox{\hbox{\strut Contribution}\hbox{\strut of growth in}\hbox{\strut output}} & \vbox{\hbox{\strut Contribution}\hbox{\strut of growth in}\hbox{\strut income similarity}} &\vbox{\hbox{\strut Contribution}\hbox{\strut of change in}\hbox{\strut trade cost measure}} & \vbox{\hbox{\strut Contribution}\hbox{\strut of change in}\hbox{\strut multilateral factors}} & \vbox{\hbox{\strut Average growth}\hbox{\strut of bilateral}\hbox{\strut trade flows}}\\
\hline
&\multicolumn{5}{|c|}{JMN 2011, unweighted}\\
\hline
& 195 & 3 & 251 & -14 & 435\\
\hline
    &\multicolumn{5}{|c|}{JMN 2011, GDP-weighted}\\
\hline
& \textbf{225} & \textbf{-11} & \textbf{290} & \textbf{-18} & \textbf{486}\\
\hline
 &  &  &  &  & \\
\hline
&\multicolumn{5}{|c|}{JMN by country, unweighted}\\
\hline
France & 154 & 24 & 168 & -17 & 329\\
\hline
UK & 168 & 18 & 122 & -18 & 290\\
\hline
USA & 312 & -51 & 322 & -15 & 568\\
\hline
Average & 222 & -7 & 263 & -17 & 461\\
\hline
 &   &  &  &  & \\
\hline
&\multicolumn{5}{|c|}{JMN by country, GDP-weighted} \\
\hline
France & 187 & -1 & 151 & -12 & 325\\
\hline
UK & 194 & 3 & 102 & -13 & 286\\
\hline
USA & 286 & -29 & 244 & -11 & 490\\
\hline
Average & 228 & -9 & 295 & -13 & 501\\
\hline
 &  &  &  &  & \\
\hline
&\multicolumn{5}{|c|}{3 countries model}\\
\hline
France & 205 & -30 & 84 & -8 & 251\\
\hline
UK & 204 & -18 & 92 & -9 & 269\\
\hline
USA & 201 & 48 & 123 & -8 & 364\\
\hline
\vbox{\hbox{\strut Unweighted}\hbox{\strut average}}  & 206 & 3 & 128 & -12 & 325\\
\hline
\vbox{\hbox{\strut GDP-weighted}\hbox{\strut average}}  & \textbf{203} & \textbf{8} & \textbf{163} & \textbf{-9} & \textbf{365}\\
\hline
\end{tabular}
\caption{Decomposition of the growth in international trade
(logarithms) with ad hoc averages and a microfounded
aggregation method. First wave of globalization,
1870-1913.  \emph{JMN 2011} refers to the averaging over dyads,
\emph{JMN by country} by country refers to the averaging over
trading partners for one country, \emph{3 countries model}
refers to the aggregation method we offer.}\label{glob1}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|c|c|c|c|c|}
\hline
\vbox{\hbox{\strut 1950-2000}\hbox{\strut }\hbox{\strut}} & \vbox{\hbox{\strut Contribution}\hbox{\strut of growth in}\hbox{\strut output}} & \vbox{\hbox{\strut Contribution}\hbox{\strut of growth in}\hbox{\strut income similarity}} &\vbox{\hbox{\strut Contribution}\hbox{\strut of change in}\hbox{\strut trade cost measure}} & \vbox{\hbox{\strut Contribution}\hbox{\strut of change in}\hbox{\strut multilateral factors}} & \vbox{\hbox{\strut Average growth}\hbox{\strut of bilateral}\hbox{\strut trade flows}}\\
\hline
&\multicolumn{5}{|c|}{JMN 2011, unweighted}\\
\hline
 & 353 & 8 & 148 & -25 & 484\\
\hline
&\multicolumn{5}{|c|}{JMN 2011, GDP-weighted}\\
\hline
 & \textbf{350} & \textbf{3} & \textbf{137} & \textbf{-17} & \textbf{473}\\
\hline
 &  &  &  &  & \\
\hline
&\multicolumn{5}{|c|}{JMN by country, unweighted}\\
\hline
France & 355 & -1 & 201 & -29 & 526\\
\hline
UK & 280 & 29 & 54 & -25 & 338\\
\hline
USA & 343 & 11 & 112 & -21 & 445\\
\hline
Average & 349 & 9 & 124 & -23 & 459\\
\hline
 &  &  &  &  & \\
\hline
&\multicolumn{5}{|c|}{JMN by country, GDP-weighted}\\
\hline
France & 365 & 2 & 271 & -21 & 617\\
\hline
UK & 319 & 1 & 202 & -20 & 502\\
\hline
USA & 349 & 27 & 165 & -16 & 525\\
\hline
Average & 361 & 21 & 190 & -19 & 553\\
\hline
 &  &  &  &  & \\
\hline
&\multicolumn{5}{|c|}{3 countries model} \\
\hline
France & 371 & -12 & 259 & -16 & 602\\
\hline
UK & 372 & -59 & 122 & -15 & 420\\
\hline
USA & 372 & -7 & 195 & -12 & 548\\
\hline
\vbox{\hbox{\strut Unweighted}\hbox{\strut average}}  & 343 & 11 & 116 & -18 & 452\\
\hline
\vbox{\hbox{\strut GDP-weighted}\hbox{\strut average}}  & \textbf{361} & \textbf{15} & \textbf{183} & \textbf{-14} & \textbf{545}\\
\hline
\end{tabular}
\caption{Decomposition of the growth in international trade
(logarithms) with ad hoc averages and a microfounded
aggregation method. Second wave of globalization, 1950-2000.}\label{glob2}
\end{table}

Tables \ref{glob1} and \ref{glob2} highlight how much the decomposition
between the decrease in trade costs and the income growth depends on the aggregation technique. When we use the microfounded aggregation method,\footnote{We selected the (GDP-weighted) average of our measure in order
to allow a clear comparison with Jacks et al.'s results. There
is of course no rational for such a summary statistic, but the
results for France, the UK and the USA assure that the main
conclusion of this exercice does not depend on averaging over
countries.} 
growth in output is the main driving force behind both waves of growth in
international trade, contributing to about 56\% (203/365) in
the first wave and 66\% (361/545) in the second one. It is no coïncidence
accident if these numbers look very much like those of table \ref{OR}: 60\% (71/107) in the first globalization and 74\% (183/246) in the second one.
Insofar as a measure of
trade costs is defined residually as everything that explains
differences between domestic and international trade flows, it
is bound to reword the information contained in an openness
ratio.

@@

Our explanation of why Jacks et al.'s aggregate measure of
trade costs yields different conclusion 
from table \ref{OR} is because it is sensitive to structure
effects. To explain this idea, and for clarity purposes, let us consider
a word where partners are symmetric (domestic trade is equal in $i$ and $j$ and $x_ij$=$x_ji$). From equation \ref{keyeq}, we have:

@@

Noting $C_{ki}$ the consumption by country $i$ of good from
region $k$, $\sigma$ the elasticity of substitution, and
$\beta_k$ a positive distribution parameter, preferences of
countries $i$ and $j$'s representative consumers are given by
their respective utility functions:

>>>>

\author[1]{Guillaume Corlay}
\author[2]{Stéphane Dupraz}
\author[3]{Claire Labonne}
\author[1]{Anne Muller}
\author[4]{Céline Antonin}
\author[5]{Guillaume Daudin\footnote{Corresponding author: guillaume.daudin@dauphine.fr}
\affil[1] {ENSAE, French National School of Statistics and Economic Administration, 3, avenue Pierre Larousse, 92245 Malakoff, FRANCE}
\affil[2] {Columbia University, 10025 New York NY, USA}
\affil[3] { Paris School of Economics / Université Paris 1 Panthéon Sorbonne -- ACPR - Banque de France, 75009, PARIS, FRANCE}
\affil[4] {Sciences Po, Observatoire Français des Conjonctures Économiques (OFCE), 75007, PARIS, FRANCE}
\affil[5] {Université Paris-Dauphine, PSL Research University, LEDa, 75016 PARIS, FRANCE
Université Paris-Dauphine, PSL Research University, LEDa, UMR [225], DIAL, 75016 PARIS, FRANCE

@@

@@

\begin{abstract}
\cite{JMN2011} offer an alternative to price gaps to quantify
trade costs. Implementing a method which consists in deducing international
trade costs from trade flows, they argue that the reduction in
trade costs was the main driving force of trade growth
during the first globalization (1870-1913), whereas economic expansion was the main driving force during the second globalization (1950-2000). We argue
that this important result is  driven by the use of an \textit{ad hoc} aggregation
method. What \cite{JMN2011} capture is the 
difference in the relative starting trade of dyads experiencing faster trade growth in the first and second globalization.
More generally, we cast doubts on the possibility to reach
conclusions of such nature with a method that infers trade
costs from trade flows, and then uses these costs to explain
trade flows. We argue that it can only rephrase the information already
contained in openness ratios.
\end{abstract}

@@

Jacks and his coauthors offer in several papers  a method for
inferring international trade costs from trade flows.\footnote{The method is developed in \cite{JMN2008}, \cite{JMN2010}, \cite{NOVY} and \cite{JMN2011}. We will use this latter paper as a reference} Using the general equilibrium model of
\cite{AW2003}, they calculate trade costs (defined as all barriers to trade, notably transportation and transaction costs) and their
evolution during the first and second waves of globalization
(1870-1913 and 1950-2000) as well as the interwar period
(1921-1939) thanks to the impressive set of data they collected
on trade flows, GDP and exports between 27
countries\footnote{Argentina, Australia, Austria, Belgium,
Brazil, Canada, Denmark, France, Germany, Greece, India,
Indonesia, Italy, Japan, Mexico, the Netherlands, New Zealand,
Norway, the Philippines, Portugal, Spain, Sri Lanka, Sweden,
Switzerland, the United Kingdom, the United States, and
Uruguay. The data contain 130 country pairs.}. They provide a
decomposition of the growth of trade caused by the reduction in
trade costs and economic expansion. They use their computations to underline
a difference of nature between the two globalizations:

@@

This is potentially an important result that sheds light on the globalization processes. However, that result is actually driven by an \textit{ad hoc}
method of aggregation that captures structure effects. The inferred difference between the two globalization comes from the fact that the dyads with the 
fastest growing trade in the first wave of globalization start with very small 
trade. This is not the case in the second wave of globalization.
Using a three-countries version of \cite{AW2003}'s model for
the aggregation of bilateral trade costs shows that there was no 
difference in nature between the two globalizations. More generally, we
cast doubt on the possibility to distinguish the effect of trade costs and economic expansion through an approach that relies only on the study of trade
flows. Deducing trade costs from trade flows, and then using
trade costs to explain trade flows is essentially a circular
reasoning. Therefore, Jacks and al.'s approach cannot be an
alternative to traditional investigations of impediments to
trade, such as commodity price gaps.

This paper starts by presenting Jacks and al.'s approach of the
measure of trade costs, and insists on its relevance at the
bilateral level to control for multilateral trade barriers in
gravity regressions (section \ref{1}). It then highlights that the
result on a difference of nature between the two globalizations
is paradoxical since it cannot be deduced from a comparison of the
evolution of openness ratios. Section \ref{2} shows that the
conclusion is only driven by the authors' \textit{ad hoc} aggregation
method. We propose a microfounded way to aggregate trade costs
and the puzzle fades away. Section \ref{3} explores the reasons
why Jacks and al.'s aggregation technique ends up providing
different results for the two globalizations. We argue that
what Jacks and al. misleadingly attribute to unequal trade costs
decreases between the two globalizations is instead a
difference in the distribution of trade growth over trading
partners.

@@

\cite{AW2003} use this model to microfound gravity equations
and solve \cite{MAC}'s border puzzle by highlighting that
bilateral trade does not depend on bilateral trade barriers \textit{per
se}, but bilateral trade barriers relative to impediments to
trade with all other trading partners. \cite{AW2003} show that the
equilibrium imposes the following relation:

@@

where $x_{ij}$ are real exports from $i$ to $j$, $y_i$ is real
output of country $i$, $y^W$ is the world real output, $\sigma$
is the elasticity of substitution, $P_i$ is the price index in
country $i$ and can be interpreted as multilateral trade
barriers, and $t_{ij}$ is the trade costs factor between $i$
and $j$. Trade costs factors are assumed to be symmetric, i.e.
$t_{ij}$ = $t_{ji}$.

Jacks and al. first depart from \cite{AW2003} by eliminating the
multilateral resistance variable ($P_i P_j$) from the gravity equation.
Following \cite{NOVY}, bilateral trade barriers are not
considered relative to multilateral trade barriers modelized by
the price index, but to domestic trade costs. In this case,
trade flows are no longer compared to outputs, but to internal
trade $x_{ii}$\footnote{Due to data limitations, the authors
use the relation $x_{ii}=GDP_i-EXPORTS_i$ to get internal
trade. We follow them. Concerns about the fact that GDP is
measured in added-value and exportations as gross value are
addressed in appendix B of \cite{JMN2011}.}. The equation above
becomes:

@@

The last equality defines $\tau_{ij}$, the geometric mean
of trade costs from country $i$ to $j$ and $j$ to $i$
\emph{relative to intra-national trade costs in countries $i$
and $j$} (no assumption of symmetry of bilateral trade costs is
imposed). It is the trade cost measure used by the authors.

This simple step offers a significant breakthrough in the
microfoundations of gravity equations. Comparing bilateral
trade flows to intranational trade allows to integrate
\cite{AW2003}'s caveat against omitted variable bias while
getting rid of multilateral trade barriers. This is important
when multilateral trade barriers cannot be estimated, e.g. when
we only have data on trade for one country. \cite{JMN2011} take
advantage of this feature in the fifth section of their article in
regressing their measure of trade costs between two countries
on a set of proxies for trade costs, such as the distance
between trade partners, tariffs, or the volatility of the
exchange rate. This is a very useful approach.\\

@@

Our point is that a method that uses trade flows
alone can teach us much about their determinants. When
departing from \cite{AW2003}'s multilateral resistances, the
arbitrage condition Jacks and al. base their computation on is an
equality between a Marginal Rate of Substitution (MRS) and a
price ratio for a CES utility function, or more precisely the
product of two such equalities, one from the program of each
trade partner. Appendix \ref{A1} derives equation (\ref{keyeq})
this way without passing through multilateral resistance. The
inference of trade costs from trade flows is tantamount to
using the relation between quantities and prices given by a
demand curve derived from fully specified exogenous
preferences.

This is of course no reason in itself to disregard the method.
Yet it cannot be used to answer \emph{``the central
question of what drives trade booms and busts''}(p. 186) as the
authors do in section 6 of their article, because the reasoning is
circular: first calculate trade costs from trade flows, then
explain trade flows from the inferred trade costs.

Formally, the authors take the logarithm of the key equation of
their article (\ref{keyeq}) to decompose the product of
bilateral trade flows between $i$ and $j$ in four terms:

@@

Since the trade costs measure is assimilated to trade flows
relative to domestic activity, they simply restate
 an information already contained in any measure of
trade flows relative to economic size, like an openness ratio
defined for instance as the ratio of exportations to GDP.
However the main conclusion in Jacks and al.'s article, namely
that the first globalization was driven more by the decrease in
trade costs than the second one, requires the absence of circularity.
Indeed, when we decompose the level of
exportations of a country as the product of its GDP and its
openness ratio, we find results that are different from theirs (see table \ref{OR}).\footnote{We follow
the authors in interpreting log differences as percentages. However, one should keep in mind that given the
size of the changes, this is a very inexact approximation.} The increase in exportations is mainly explained by GDP growth for the large majority of the 27 countries in the
data. On average, this decomposition attributes 74\% (183/246) of
the growth in trade to the increase in GDP in the second
globalization and 60\% in the first globalization. This is very much in contrast to 
Jacks and al.\footnote{Contributions of growth in income
similarity and of change in multilateral factors are negative.} :

@@

This contrast between both decompositions of the growth of exports is at the center of the argument by Jack et al. As the next section shows, it is actually driven by the \textit{ad hoc} way they aggregate trade costs.

\section{\label{2} \textit{Ad hoc} aggregation of trade costs}

Jacks and al.'s conclusion on a difference of nature between the
two globalizations is based on an aggregate trade costs
measure that provides a summary statistic of the
evolution of trade costs across all dyads in the sample, as
well as for different regions of the world. To move from
bilateral costs $\tau_{ij}$ to an aggregate measure of trade
costs, the authors use an arithmetic mean over dyads, weighted
by the sum of the GDP of the two trade partners.

@@

\cite{AW2003}'s model can be used to calculate the aggregate
trade costs a single country faces in its trade with all
its trading partners. Indeed, using a two-countries model where
all trade partners of country $i$ are treated as one single
country provides an aggregate measure of trade costs that does
not rely on any \textit{ad hoc} aggregation method but instead on
microfoundations. Since all trade flows data are not available,
it is also possible to use a three-countries model, with
country $i$, its trading partners present in the data, and the
rest of the world.

Let us note $\upsilon_{i}$ the corresponding trade cost faced
by country $i$ with all its trading partners in the data. Its
expression is given by Jacks and al.'s key equation
(\ref{keyeq}) applied to a three-countries model:

@@

Using this measure of trade costs, we provide the same
decomposition in four terms of the increase in trade flows
during the two globalizations as in the authors' article for
the countries with the most trade partners available in the
data (France (24), the UK (25), the USA (23)). We also provide
the unweighted and end-of-period-GDP-weighted averages for all
countries in the sample. Results are displayed in tables
\ref{glob1} and \ref{glob2} along with the results with the
Jacks and al's method.\footnote{They include unweighted and
end-of-period-GPD-weighted averages over country dyads of
bilateral trade costs, such as provided in \cite{JMN2011}. We
also display for France, the UK and the USA the trade cost
measure averaged over trading partners faced by one country, as
the results are presented (with a smaller data set) in
\cite{JMN2008, JMN2010}. We also provide both averages of this
measure over all countries in the data set.}\\

\begin{table}
\centering
\begin{tabular}{|l|c|c|c|c|c|}
\hline
\vbox{\hbox{\strut 1870-1913}\hbox{\strut }\hbox{\strut}} & \vbox{\hbox{\strut Contribution}\hbox{\strut of growth in}\hbox{\strut output}} & \vbox{\hbox{\strut Contribution}\hbox{\strut of growth in}\hbox{\strut income similarity}} &\vbox{\hbox{\strut Contribution}\hbox{\strut of change in}\hbox{\strut trade cost measure}} & \vbox{\hbox{\strut Contribution}\hbox{\strut of change in}\hbox{\strut multilateral factors}} & \vbox{\hbox{\strut Average growth}\hbox{\strut of bilateral}\hbox{\strut trade flows}}\\
\hline
&\multicolumn{5}{|c|}{JMN 2011, unweighted}\\
\hline
& 195 & 3 & 251 & -14 & 435\\
\hline
    &\multicolumn{5}{|c|}{JMN 2011, GDP-weighted}\\
\hline
& \textbf{225} & \textbf{-11} & \textbf{290} & \textbf{-18} & \textbf{486}\\
\hline
 &  &  &  &  & \\
\hline
&\multicolumn{5}{|c|}{JMN by country, unweighted}\\
\hline
France & 154 & 24 & 168 & -17 & 329\\
\hline
UK & 168 & 18 & 122 & -18 & 290\\
\hline
USA & 312 & -51 & 322 & -15 & 568\\
\hline
Average & 222 & -7 & 263 & -17 & 461\\
\hline
 &   &  &  &  & \\
\hline
&\multicolumn{5}{|c|}{JMN by country, GDP-weighted} \\
\hline
France & 187 & -1 & 151 & -12 & 325\\
\hline
UK & 194 & 3 & 102 & -13 & 286\\
\hline
USA & 286 & -29 & 244 & -11 & 490\\
\hline
Average & 228 & -9 & 295 & -13 & 501\\
\hline
 &  &  &  &  & \\
\hline
&\multicolumn{5}{|c|}{3 countries model}\\
\hline
France & 205 & -30 & 84 & -8 & 251\\
\hline
UK & 204 & -18 & 92 & -9 & 269\\
\hline
USA & 201 & 48 & 123 & -8 & 364\\
\hline
\vbox{\hbox{\strut Unweighted}\hbox{\strut average}}  & 206 & 3 & 128 & -12 & 325\\
\hline
\vbox{\hbox{\strut GDP-weighted}\hbox{\strut average}}  & \textbf{203} & \textbf{8} & \textbf{163} & \textbf{-9} & \textbf{365}\\
\hline
\end{tabular}
\caption{Decomposition of the growth in international trade
(logarithms) with \textit{ad hoc} averages and a microfounded
aggregation method. First wave of globalization,
1870-1913.  \emph{JMN 2011} refers to the averaging over dyads,
\emph{JMN by country} by country refers to the averaging over
trading partners for one country, \emph{3 countries model}
refers to the aggregation method we offer.}\label{glob1}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|c|c|c|c|c|}
\hline
\vbox{\hbox{\strut 1950-2000}\hbox{\strut }\hbox{\strut}} & \vbox{\hbox{\strut Contribution}\hbox{\strut of growth in}\hbox{\strut output}} & \vbox{\hbox{\strut Contribution}\hbox{\strut of growth in}\hbox{\strut income similarity}} &\vbox{\hbox{\strut Contribution}\hbox{\strut of change in}\hbox{\strut trade cost measure}} & \vbox{\hbox{\strut Contribution}\hbox{\strut of change in}\hbox{\strut multilateral factors}} & \vbox{\hbox{\strut Average growth}\hbox{\strut of bilateral}\hbox{\strut trade flows}}\\
\hline
&\multicolumn{5}{|c|}{JMN 2011, unweighted}\\
\hline
 & 353 & 8 & 148 & -25 & 484\\
\hline
&\multicolumn{5}{|c|}{JMN 2011, GDP-weighted}\\
\hline
 & \textbf{350} & \textbf{3} & \textbf{137} & \textbf{-17} & \textbf{473}\\
\hline
 &  &  &  &  & \\
\hline
&\multicolumn{5}{|c|}{JMN by country, unweighted}\\
\hline
France & 355 & -1 & 201 & -29 & 526\\
\hline
UK & 280 & 29 & 54 & -25 & 338\\
\hline
USA & 343 & 11 & 112 & -21 & 445\\
\hline
Average & 349 & 9 & 124 & -23 & 459\\
\hline
 &  &  &  &  & \\
\hline
&\multicolumn{5}{|c|}{JMN by country, GDP-weighted}\\
\hline
France & 365 & 2 & 271 & -21 & 617\\
\hline
UK & 319 & 1 & 202 & -20 & 502\\
\hline
USA & 349 & 27 & 165 & -16 & 525\\
\hline
Average & 361 & 21 & 190 & -19 & 553\\
\hline
 &  &  &  &  & \\
\hline
&\multicolumn{5}{|c|}{3 countries model} \\
\hline
France & 371 & -12 & 259 & -16 & 602\\
\hline
UK & 372 & -59 & 122 & -15 & 420\\
\hline
USA & 372 & -7 & 195 & -12 & 548\\
\hline
\vbox{\hbox{\strut Unweighted}\hbox{\strut average}}  & 343 & 11 & 116 & -18 & 452\\
\hline
\vbox{\hbox{\strut GDP-weighted}\hbox{\strut average}}  & \textbf{361} & \textbf{15} & \textbf{183} & \textbf{-14} & \textbf{545}\\
\hline
\end{tabular}
\caption{Decomposition of the growth in international trade
(logarithms) with \textit{ad hoc} averages and a microfounded
aggregation method. Second wave of globalization, 1950-2000.}\label{glob2}
\end{table}

Tables \ref{glob1} and \ref{glob2} highlight how much the decomposition
between the decrease in trade costs and the income growth depends on the aggregation technique. When we use the microfounded aggregation method,\footnote{We selected the (GDP-weighted) average of our measure in order
to allow a clear comparison with Jacks and al.'s results. There
is of course no rational for such a summary statistic, but the
results for France, the UK and the USA assure that the main
conclusion of this exercice does not depend on averaging over
countries.} 
growth in output is the main driving force behind both waves of growth in
international trade, contributing to about 56\% (203/365) in
the first wave and 66\% (361/545) in the second one. It is no coincidence
if these results are very similar to the ones in table \ref{OR}: 60\% (71/107) in the first globalization and 74\% (183/246) in the second one.
Insofar as a measure of
trade costs is defined residually as everything that explains
differences between domestic and international trade flows, it
is bound to reword the information contained in an openness
ratio.

@@

Our explanation of why Jacks and al.'s aggregate measure of
trade costs yields different conclusion 
from table \ref{OR} is because it is sensitive to structure
effects. To explain this idea, and for clarity purposes, let us consider
a word where partners are symmetric (domestic trade is equal in $i$ and $j$ and $x_ij$=$x_ji$). From equation \ref{keyeq}, we have:

@@

Let us note $C_{ki}$ the consumption by country $i$ of good from
region $k$, $\sigma$ the elasticity of substitution, and
$\beta_k$ a positive distribution parameter, preferences of
countries $i$ and $j$'s representative consumers are given by
their respective utility functions:
